Five-Dimensional Kaluza--Klein Cosmology in Lovelock--Cartan Theory

### 2016-09-10 KKosmology.tex
Version by Criz

\documentclass[aps,prd,12pt,superscriptaddress,showpacs,showkeys,longbibliography,reprint]{revtex4-1}
\usepackage{calrsfs} 
\usepackage{amsmath,amsthm,latexsym,amssymb,amsfonts}
\usepackage{xcolor}
\usepackage{textcomp}
\usepackage[%
  colorlinks=true,
  urlcolor=blue,
  linkcolor=blue,
  citecolor=blue
]{hyperref}
\usepackage{etoolbox}
\usepackage{breqn}

@@

\title{Five Dimensional Kaluza--Klein Cosmology in Lovelock--Cartan Theory}

@@

\author{Cristobal \surname{Corral}}
\affiliation{\UTFSM.}
\affiliation{\CCTVal.}

@@

%% --------- Abstract
\begin{abstract}
 We study the Kaluza-Klein dimensional reduction of an $N$ dimensional Riemann--Cartan manifold when the compact section is $S^1$. First, we focus on the details of dimensional reduction in the presence of torsion to then study the particular Einstein--Cartan theory with the Gauss--Bonnet term in five dimensions. We look for the general cosmological solution of the Friedman-Robertson-Walker familly in the four-manifold to interpret the remaining fields. The model, treated in five dimensional vacuum, presents torsion both in the reduced manifold as well as in the compact section, manifest as a non-vanishing four dimensional matter field that adds to the effective energy density and pressure. This allows the model to present expanding periods as well as contracting universes, and a radius of the compact section that it asymtotes zero in some solutions. 
\end{abstract}

@@

\section{Introduction}
The geometric theory of gravity, General Relativity (GR), has been highly successful in describing the physics at large scales. Two major predictions were recently confirmed in one experiment that invoque two of the most fascinating conepts that this theory developed: the propagation of ripples of spacetime, gravitational waves, originated by the merge of two black holes where detected by the LIGO collaboration \cite{Abbott:2016blz} confirming the existence of both, astrophysical balck holes as well as the unprecedented nature of the signal detected. With this, we can rely in the geometric nature of this interaction. At the galactic scale, the gravitational lensing produced by the local distribution of energy and matter suggest the presence of a kind of matter that is not seen through telescopes. This abundance of matter in the galaxy is also compatible with the velocity profyle of stars at the outer regions of it and is needed to reach the current state of structures formation. This matter would interact mostly, if not only, through gravity \cite{Sofue:2000jx}. At the cosmic scale, observatinal evidence points out an accelerated current state of the Universe \cite{Riess:1998cb}. Such a behaviour needs the existence of a form of vacuum energy of exotic nature to give account of the accelerated expansion. This is evidence of unknown physics, and it has been the motivation for looking for new gravitational degrees of freedom beyond Einsten's GR. Among the familly of extended gravity theories, higher dimensional gravity can give us a clue to explore the manifestation of new degrees of freedom as material fields of certain class. Along this lines is that we consider here an extended model of gravity given by the so called Lovelock action~\cite{Lovelock:1971yv}. 
The Einstein--Hilbert action with cosmological constant is the unique action in four dimensions that yields second order field equations for the metric. In higher dimensions, however, this does not remains true. A particular combination of squared terms in the curvature can be added to the gravitational action to give rise a second order system for the metric with no ghosts. The particular quadratic term is called Gauss--Bonnet term, and in $N$ dimensions reads
\begin{equation}\label{GB}
  \mathcal{L}_{GB}=\text{d}^N \!x \,\sqrt{-g}\left(\tilde{R}^2-4\tilde{R}_{\mu\nu}\tilde{R}^{\mu\nu}+\tilde{R}_{\alpha\beta\mu\nu}
  \tilde{R}^{\alpha\beta\mu\nu}\right),
\end{equation}
where $\tilde{R}_{\alpha\beta\mu\nu}$ is the Riemannian curvature of a manifold with metric $g_{\mu\nu}$ and $g$ its determinant. $\tilde{R}_{\mu\nu}$ and $\tilde{R}$ are the Ricci tensor and Ricci scalar built with the curvature.

Written in four dimensions, Eq.~\eqref{GB} adds no dynamics to the metric, since it represents a topological invariant proportional to the Euler characteristic and therefore can always be written locally as a boundary term, while in lower dimensions vanishes identically. Is then natural to extend gravitational Lagrangians by dimensional continuation of Euler densities to build a general theory for a metric in arbitrary dimensions, while keeping the two important features of Einstein's General Relativity just mentioned: the absence of ghosts and second order field equations. These family of Lagrangians for gravity where introduced by Lovelock and bear his name. It can be built in arbitrary dimensions as a linear combination of dimensionally continued topological invariants. Equivalently, the previous Lagrangian can be written in exterior forms, that provide shorter expressions together with a natural framework for further generalizations, in $N$ dimensions reads
\begin{equation}\label{GB2}
\mathcal{L}_{GB} = \epsilon_{a_1...a_N} \tilde{R}^{a_1a_2}\wedge\tilde{R}^{a_3a_4}\wedge e^{a_5}\wedge...\wedge e^{a_N}.
\end{equation}
This Lagrangian was also identified as the low energy correction for a spin two field in string theory~\cite{Zwiebach:1985uq}. Thus, the consideration of further Lovelock terms in a higher dimensional manifold can be interpreted as semiclassical corrections of a gravity action.
The higher dimensional scenario seems to be a natural framework for unification. Early in the last century, it was realized by T. Kaluza and developed by O. Klein, that the existence of one extra direction in Einstein geometric theory of gravity would give rise a unified picture of gravity and electromagnetism together with a spectrum of new heavy particles~\cite{Kaluza:1921tu,*Klein:1926tv}. This opened the posibility of a geometrical understanding of interactions given by non-Abelian gauge groups as a concequence of the topology of a spatial compact section in a higher dimensional spacetime. Eventhough the original model, and further developments have not yet reached a realistic phenomenology, it is clear that this possibility of unification of internal and spacetimes symmetries, together with a material manifestation of geometric degrees of freedom provides a fruitful arena for looking for clues of unity among the interactions. The compactification of higher Lovelock terms has been considered \cite{MuellerHoissen:1985mm,MuellerHoissen:1989yv} and its consequences in cosmology in \cite{MuellerHoissen:1985ij,Deruelle:1986iv,Deruelle:1989fj}. In the Riemannian five dimensional theory, exact wormholes solutions have been found in vacuum \cite{Dotti:2006cp,Dotti:2007az}, and suported by non-exotic matter, making the sign of the Gauss-Bonnet coupling the source of exotic behaviour \cite{Mehdizadeh:2015jra}, higher order Lovelock models \cite{Mehdizadeh:2015dta}, as well as in higher order compactified Lovelock theories with torsion \cite{Canfora:2008ka} also have these solutions. Classically, wormholes must be supported by a kind of phantom energy, compatible with the cosmological hypothesis. We see their theoretical existence in these extended models as a good omen for the search of new degrees of freedom that could provide explanation for the exotic matter-energy abundance in the Universe. \\
Torsional degrees of freedom, on the other hand, as a further feature of spacetime had a \textit{renaissance} with the works of D. W. Sciama and T. Kibble \cite{Kibble:1961ba} in Poincar\'e gauge theory. In that framework, spacetime torsion arises naturally as the field strength of a gauge connection for the Poincar\'e group (see \cite{Hehl:1976kj,Blagojevic:2002du} for a complete review). The pose of gravity as a gauge theory has major benefits in the problem of a unifyied picture for interactions since it allows it to interpret it as arising by means of a localization procedure, where spacetime is a Riemann-Cartan manifold. In this context, it has been shown that torsional functions produce an accelerating profyle compatible with the current state of the Universe \cite{Shie:2008ms}, as well as non-minimal coupling of Euler forms in Riemann--Cartan spacetimes allow torsion to behave as matter of exotic characteristics \cite{Toloza:2013wi} customary for expansion.
It becomes essential to understand Kaluza--Klein's ideas in the presence of torsion. Work on the subject reports the phenomenology of extra dimensional torsion \cite{Kalinowski:1980da}, metric dependent torsional models of extra dimensions \cite{Shankar:2012vd} and its consequences in cosmology \cite{Chen:2009ep}, compatification of higher dimensional Brans--Dicke models with torsion \cite{German:1993bq}. Torsion free solutions where found in \cite{Aros:2007nn} for models in first order compactified gravity.\\
Eventhough at the macroscopic scale, spacetime seems to lack torsion, there is no logical constraint to assume it non-existing, as E. Cartan argued, moreover since vacuum predictions of Einstein--Cartan theory hold the experimental tests of General Relativity.\\
This work is organized as follows: in Sec.~\ref{KK} we discuss the introduction of torsional degrees of freedom compatible with the Kaluza--Klein geometric reasoning, as well as fix our notation and conventions. Sec.~\ref{5EGB} studies the dynamics of the general Lovelock--Cartan action in five-dimensional spacetime and its dimensional reduction. Then, in Sec.~\ref{cosmos} we look for cosmological solutions in the reduced four-dimensional manifold of the Friedmann--Robertson--Walker class, with torsion both in the reduced manifold and as new dynamical fields arising from the compact section taken to be $S^1$. Conslusions and remarks are given next in Sec.~\ref{conclusions}. We also found appropriate to incorporate a number of appendices to make this work a self contained article.

>>>>

\documentclass[aps,prd,12pt,superscriptaddress,showpacs,showkeys,longbibliography,reprint,nofootinbib]{revtex4-1}
\usepackage{calrsfs} 
\usepackage{amsmath,amsthm,latexsym,amssymb,amsfonts}
\usepackage{xcolor}
\usepackage{textcomp}
\usepackage[%
  colorlinks=true,
  urlcolor=blue,
  linkcolor=blue,
  citecolor=blue
]{hyperref}
\usepackage{etoolbox}
\usepackage{breqn}

@@

\title{Five-Dimensional Kaluza--Klein Cosmology in Lovelock--Cartan Theory}

@@

\author{Cristobal \surname{Corral}}
\email{cristobal.corral@usm.cl}
\affiliation{\UTFSM.}
\affiliation{\CCTVal.}

@@

%% --------- Abstract
\begin{abstract}
 We study the Kaluza--Klein dimensional reduction of the Lovelock--Cartan theory on a $N$-dimensional Riemann--Cartan manifold, where the compact manifold has $S^1$ topology. First, we focus on the dimensional reduction in presence of torsion, to proceed with the Lovelock--Cartan theory in five dimensions. We find cosmological solutions of the Friedman--Robertson--Walker class for the dimensionally reduced theory. In the five-dimensional vacuum, this theory allows a nontrivial torsion in contrast to the Einstein--Cartan theory. After the dimensional reduction is performed, the torsional degrees of freedom induce a nonvanishing energy-momentum tensor in four dimensions, which behaves as a perfect fluid. Cosmological solutions describing accelerated expansion phases as well as contracting solutions are found. The model shows a dynamical compactification of the extra dimension in some regions of the parameter space. 
\end{abstract}

@@

\section{Introduction}

%The theory of General Relativity (GR) has been proven to be one of the most successful description of gravity nowadays. In this framework, the gravitational interaction is understood as the geometrical manifestation of the spacetime. 
The experimental status of General Relativity (GR), regarding the solar system tests~\cite{Will:2014kxa} and the detection of gravitational waves by the LIGO collaboration~\cite{Abbott:2016blz}, have settled GR as one of the most accepted theories of gravity so far. However, despite these goals, the difficulties on finding out a quantum theory of gravitation, has driven the community to think that GR is not the ultimate gravitational theory.

At the galactic scale, on the one hand, the gravitational lensing produced by the local distribution of energy and matter, suggests the presence of an exotic form of matter, which have not been seen by the current light-based telescopes~\cite{Sofue:2000jx}. Such an abundance in the galaxy, is also compatible with the velocity profile of the stars at the outter regions of it and is needed to reach the current state of structures formation. This matter would interact mostly (if not only) through gravity~\cite{Sofue:2000jx}. At the cosmic scale, on the other hand, experimental data obtained from the type Ia supernovae observations, indicates that our Universe is passing through a phase of accelerated expansion~\cite{Riess:1998cb}. The latter experimental fact suggests the existence of an exotic form of energy in order to generate such an observed behaviour.

The shortcomings of GR on describing these manifestations of new physics, have been taken as the main motivation to look for new gravitational degrees of freedom beyond GR. Among the families of possible extensions, higher dimensional gravity could shed some lights on the nature of these new degrees of freedom, appearing as material fields within the dimensionally reduced theory. For instance, as it was shown in the early works of T. Kaluza and O. Klein, the existence of one extra dimension within GR would give rise to an unified picture of gravity and electromagnetism, along with a spectrum of new heavy particles~\cite{Kaluza:1921tu,*Klein:1926tv}. This fact opened the posibility of a geometrical understanding of interactions, where the gauge group arise as a consequence of the topology of the spatial compact manifold in a higher dimensional spacetime, while the gauge fields arise through the dimensional reduction. Even though the original model have not reached a realistic phenomenology so far, the idea of higher dimensions has been taken into account by string theory~\cite{Green:1987sp,*Green:1987mn} and by ADD~\cite{ArkaniHamed:1998rs,*Antoniadis:1998ig} and RS~\cite{Randall:1999ee,*Randall:1999vf} models, as an attempt of unification and as a solution to the hierarchy problem, respectively.

It is well-known that, in four dimensions, the Einstein--Hilbert action with cosmological constant is the unique theory which leads to second order field equations for the metric. In higher dimensions, however, this is no longer true. A particular combination of squared terms in the curvature can be added to the gravitational action, giving rise to second order field equations for the metric. The most general theory in arbitrary dimensions, which preserves the same features of the four dimensional Einstein--Hilbert action with cosmological constant, is described by the Lanczos--Lovelock action~\cite{Lanczos:1938sf,Lovelock:1971yv}. Such a theory, possesses no ghosts~\cite{Zumino:1985dp} and it has the same degrees of freedom as the Einstein--Hilbert Lagrangian in arbitrary dimensions~\cite{Henneaux:1990au}. It is endowed with the so-called Gauss--Bonnet term which, in $N$-dimensions, reads as
\begin{equation}\label{GB}
  \mathcal{L}_{GB}=\text{d}^N \!x \,\sqrt{-g}\left(\tilde{R}^2-4\tilde{R}_{\mu\nu}\tilde{R}^{\mu\nu}+\tilde{R}_{\alpha\beta\mu\nu}
  \tilde{R}^{\alpha\beta\mu\nu}\right),
\end{equation}
where $\tilde{R}_{\alpha\beta\mu\nu}$ is the Riemannian curvature of a manifold with metric $g_{\mu\nu}$ and $g$ its determinant. $\tilde{R}_{\mu\nu}$ and $\tilde{R}$ are the Ricci tensor and Ricci scalar, respectively, built up by means of the Riemannian curvature.

In four dimensions, the Gauss--Bonnet term~\eqref{GB} adds no dynamics to the metric, since it represents a topological invariant proportional to the Euler characteristic class, which can be written locally as a boundary term. In dimensions higher than four, however, the Gauss--Bonnet term does contribute to the field equations, while preserving the same features of GR. Equivalently, the previous $N$-dimensional Lagrangian can be written in exterior forms as
\begin{equation}\label{GB2}
\mathcal{L}_{GB} = \epsilon_{a_1...a_N} \tilde{R}^{a_1a_2}\wedge\tilde{R}^{a_3a_4}\wedge e^{a_5}\wedge...\wedge e^{a_N}.
\end{equation}
This Lagrangian was also identified as the low energy correction for a spin two field in string theory~\cite{Zwiebach:1985uq}. Thus, the consideration of further Lanczos--Lovelock terms in a higher dimensional manifold can be interpreted as the semiclassical corrections to the gravity action.

The compactification of higher Lanczos--Lovelock terms has been considered in~\cite{MuellerHoissen:1985mm,MuellerHoissen:1989yv} and its consequences in cosmology in Refs.~\cite{MuellerHoissen:1985ij,Deruelle:1986iv,Deruelle:1989fj}. In the Riemannian five-dimensional theory, exact wormholes solutions have been found in vacuum in Refs.~\cite{Dotti:2006cp,Dotti:2007az}. The same kind of solutions were found without exotic matter, making the sign of the Gauss-Bonnet coupling the source of such an exotic behaviour~\cite{Mehdizadeh:2015jra}. These latter solutions were also obtained in higher order Lanczos--Lovelock models~\cite{Mehdizadeh:2015dta}, as well as in the higher order compactified Lanczos--Lovelock theories with torsion~\cite{Canfora:2008ka}. Classically, wormholes must be supported by a kind of phantom energy, compatible with the cosmological hypothesis. Their existence in these extended models, implies the presence of degrees of freedom that could provide an explanation for the exotic matter-energy abundance in the Universe.

On the other hand, the renaissance of the Riemann--Cartan geometries came through the advent of the pioneer works of Sciama~\cite{Sciama:1962} and Kibble~\cite{Kibble:1961ba} on Poincar\'e gauge theories (see also Refs.~\cite{Hehl:1976kj,Blagojevic:2002du}). In such a framework, the torsion appears as the field strength of translations. %In such a framework, the vielbein is interpreted as the gauge connection of translations and the torsion as its field strength. The formulation of gravity as a gauge theory has major benefits in the problem of a unified picture of interactions, since it allows to interpret it as arising through a localization procedure, where the spacetime is a Riemann--Cartan manifold. 
%Due to this, several authors have relaxed the torsion-free condition since, as E. Cartan argued, there is no a fundamental reason to assume a torsion-free condition {\it a priori}. 
Moreover, the vacuum predictions of its simplest formulation --the Einstein--Cartan theory-- holds the experimental tests of GR. The cosmological consequences of non-Riemannian geometries has been widely studied in the literature (for a review, see Ref.~\cite{Puetzfeld:2004yg}). %For instance, the pioneer works of Kopczy\'nski showed how the introduction of a spinning dust, acting as a source of the curvature and torsion, avoids the appearance of cosmological singularities~\cite{Kopczynski:1972,Kopczynski:1973}. In the context of the early evolution of the Universe, the spin density can drive the inflationary phase within the Einstein--Cartan theory, along with the avoidance of singularities~\cite{Gasperini:1986mv}. It has been shown in Ref.~\cite{Shie:2008ms} that torsional degrees of freedom might produce the current phase of accelerated expansion of the Universe. Additionally, the nonminimal coupling of the Euler characteristic form with a scalar field in Riemann--Cartan spacetimes, allows the torsional degrees of freedom to behave as an exotic induced-matter content, customary for the phase of accelerated expansion~\cite{Toloza:2013wi}.

In the framework of the Kaluza--Klein theories in higher dimensional Riemann--Cartan geometries, phenomenology of the extra dimensional torsion was found in Ref.~\cite{Kalinowski:1980da}, metric dependent torsional models of extra dimensions was studied in~\cite{Shankar:2012vd}, along with its consequences in cosmology~\cite{Chen:2009ep}. Compatification of higher dimensional Brans--Dicke models with torsion was considered in Ref.~\cite{German:1993bq}. Torsion-free black-hole solutions were found in~\cite{Aros:2007nn}, for first order compactified gravity.

In this work, we present a new class of cosmological solutions for the Lovelock--Cartan theory~\cite{Mardones:1990qc} in five dimensions, outside the Chern--Simons point. The geometry is described by the five-dimensional Riemann--Cartan manifold, where the compact manifold is $S^1$ and the four dimensional manifold is described by means of the Friedmann--Robertson--Walker metric. Such a theory implies a nonvanishing torsion in vacuum due to the presence of the Gauss--Bonnet term, in contrast to the five-dimensional Einstein--Cartan theory. After the dimensional reduction is performed, the torsional degrees of freedom adds to the energy density and pressure of the four dimensional matter fields. In some cases, the solutions avoid the appearance of initial singularities and drive the accelerated expansion of the Universe, while the radius of the compact manifold goes to zero.

This work is organized as follows: In Sec.~\ref{KK} we introduce the torsional degrees of freedom compatible with the Kaluza--Klein geometry, as well as fixing our notation and conventions. In Sec.~\ref{5EGB}, we study the dynamics of the general Lovelock--Cartan action in five-dimensional spacetime and its dimensional reduction. In Sec.~\ref{cosmos}, we look for cosmological solutions in the reduced four-dimensional manifold of the Friedmann--Robertson--Walker class with nontrivial torsion in both: the reduced manifold as well as new dynamical fields arising from the compact manifold taken to be $S^1$. Conclusions and remarks are given in Sec.~\ref{conclusions}. We also incorporate a number of appendices to make this work a self-contained article.

Let $M_N$ be a $N$-dimensional differential manifold. Every quantity defined on $M_N$ will be denoted by hats $\hat{x}$. Capital Greek characters (coordinate indices) and capital Latin characters (Lorentz indices) run over the $N$ dimensions, while lower case run in the $(N-1)$-dimensional reduced manifold.

Let $\hat{\omega}^{AB}$ be the spin connection 1-form describing the affine structure of the $N$-dimensional manifold and $\hat{e}^A=\hat{e}^{A}_{\ \Gamma}\,\text{d}\hat{x}^\Gamma$ be the vielbein 1-form defining the metric structure of the same manifold through the relation $\hat{g}_{\Gamma\Delta}=\hat{e}^{A}_{\ \Gamma}\hat{e}^{B}_{\ \Delta}\hat{\eta}_{AB}$. The vielbein is the mapping that relates coordinate indices (Greek characters) with the Lorentz indices (Latin characters). In our convention, the flat Lorentzian metric will be $\hat{\eta}_{AB}=\,$diag$\,(-,+,...,+)$, and the $N$-dimensional Levi-Civita symbol $\hat{\epsilon}_{A_1...A_N}$ is reduced by defining
\begin{align*}
\hat{\epsilon}_{a_1...a_{N-1}N} \equiv \epsilon_{a_1...a_{N-1}},
\end{align*} 
since the coordinates of $M_N$ run from $0$ to $N-1$. Our convention fixes $\hat{\epsilon}_{01...N}=1$. Hereon, Riemannian fields (torsion-free) will be explicited by a tilde, as it was done in Eqs.~\eqref{GB} and~\eqref{GB2} for the curvature.

Due to the topology of the manifold, we can expand the depenence of the fields on the extra coordinate, $z$, in Fourier series as  
\begin{align}\label{Fourier}
\hat{\varrho}(\hat{x}^\Gamma)&=\sum_n\varrho_{(n)}(x)e^{in z},
\end{align}
where $x$ denotes the coordinates of the $(N-1)$-manifold, collectively. Henceforth, we will focus in the $n=0$ mode of the expasion, also refered to as the low energy sector.

The Kaluza--Klein (KK) ansatz for the metric lies on the premise that the compact manifold of $M_N$ is orthogonal to the rest of the manifold at each point. This leads to the following metric structure
\begin{equation}
  \hat{g}_{\Gamma\Delta} =
  \begin{pmatrix}
    g_{\gamma\delta} +\frac{\hat{g}_{\gamma z}\hat{g}_{\delta z}}{\hat{g}_{zz}}&\hat{g}_{\gamma z}\\
    \hat{g}_{z\delta} & \hat{g}_{zz}
  \end{pmatrix}
  =
  \begin{pmatrix}
    g_{\gamma\delta} + \phi A_\gamma A_\delta&\phi A_\gamma\\
    \phi A_{\delta} & \phi
  \end{pmatrix},
\end{equation}
and its inverse
\begin{equation}
  \hat{g}^{\Gamma\Delta}=
  \begin{pmatrix}
    g^{\gamma\delta}&-A^\gamma\\
    -A^{\delta} & \phi^{-1}+A^2
  \end{pmatrix}.
\end{equation}
It introduces a scalar field $\phi$ and a vector field $A_\mu$ as new gravitational degrees of freedom, materialized in the reduced manifold. The vielbein which holds this structure for the metric has the following form
\begin{equation}
  \label{Dvielbein}
  \hat{e}^A_{\ \ \Gamma} =
  \begin{pmatrix}
    \hat{e}^a_{\ \gamma}& 0\\
    \hat{e}^N_{\ \gamma} & \hat{e}^N_{\ \ z}
  \end{pmatrix}
  =
  \begin{pmatrix}
    e^a_{\ \gamma}& 0\\
    \sqrt{\phi}A_\gamma & \sqrt{\phi}
  \end{pmatrix},
\end{equation}
and its inverse, $\hat{E}_A^{\ \ \Gamma}$, defined such that $\hat{E}_A^{\ \ \Gamma}\hat{e}^A_{\ \ \Delta}=\delta^\Gamma_{\Delta}$, reads
\begin{equation}
  \label{Dinversevielbein}
  \hat{E}_A^{\ \ \Gamma} =
  \begin{pmatrix}
    \hat{E}_a^{\ \gamma}& 0\\
    \hat{E}_a^{\ z} & \hat{E}_N^{\ \ z}
  \end{pmatrix}
  =
  \begin{pmatrix}
    E_a^{\ \gamma}& 0\\
    -A_a & \sqrt{\phi}^{-1}
  \end{pmatrix},
\end{equation}
and it shows that the compact manifold $S^1$ has its own tangent space, $T_pS^1$, at each point $p\in M_N$, and is independent to $T_pM_{N-1}$ in the sense that they do not mix. It seems reasonable to think that if we want two manifolds to be orthogonal to one another, they will not mix vectors after the tangent mapping.

The vielbeins are defined modulo Lorentz transformations. Any transformed basis $\hat{e}^{\prime A}=\hat{\Lambda}^A_{\ B}\,\hat{e}^B$,
where $\hat{\Lambda}$ is a Lorentz matrix, is as suitable as $\hat{e}^A$, and therefore share the structure in Eq.~\eqref{Dvielbein}. The Lorentz transformations are then reduced by this fact, imposing the condition $\hat{\Lambda}^a_{\ N}=0$.

A Riemannian connection compatible with the $N$-dimensional vielbein~\eqref{Dvielbein}, is built under the premise that $\mbox{d}\hat{e}^A + \hat{\tilde{\omega}}^{AB} \wedge \hat{e}^A = 0$. Thus, we find
\begin{align}
  \label{DRiemannconnection}
  \hat{\tilde{\omega}}^{ab}&=\tilde{\omega}^{ab}-\frac{1}{2}\sqrt{\phi}F^{ab}\hat{e}^N,\\
  \hat{\tilde{\omega}}^{Na}&=\frac{1}{2}\sqrt{\phi}F^a_{\ \ l}e^l+\frac{1}{2}\partial^a\ln\phi\hat{e}^N,
\end{align}
where $\tilde{\omega}^{ab}$ is the Riemannian spin connection of the reduced manifold. The construction of the five-dimensional Einstein--Hilbert action by means of the spin connection~\eqref{DRiemannconnection}, leads to the original KK theory. After the dimensional reduction is performed, the usual $U(1)$-invariant gauge field strength, 
\begin{align*}
F=\text{d}A=\frac{1}{2}F_{ab}\, e^a\wedge e^b,
\end{align*} 
appears nonminimally coupled to the scalar field.

Nevertheless, in a Riemann--Cartan geometry $M_N$ is not entirely described by $\hat{\tilde{\omega}}^{AB}$, but by a more general spin connection, independent of the metric degrees of freedom (see Appendix \ref{Riemann-Cartan} for details). We can assume a general connection 1-form of the same type (regarding its $M_{N-1}\times S^1$ decomposition) as in Eq.~\eqref{DRiemannconnection}. Thus
the most general spin connection on $M_N$ compatible with the KK decomposition is given by 
\begin{equation}\label{Nconnection}
  \hat{\omega}^{AB} \equiv
  \begin{pmatrix}
    \omega^{ab}+\alpha^{ab}\hat{e}^N&\beta^a+\gamma^a\hat{e}^N\\
    -\beta^b-\gamma^b\hat{e}^N&0
  \end{pmatrix},
\end{equation}
which contemplates the torsional degrees of freedom appart from the usual metric ones.

The decomposition \eqref{Nconnection} adds new metric-independent fields into the reduced theory, after the dimensional reduction is performed. For instance, $\alpha^{ab}$ is a rank-2 Lorentz-valued antisymmetric $0$-form, which adds a new spin-$1$ field into the matter content of the reduced theory. The $\beta^a$ piece of the extra dimensional spin connection, is a Lorentz-valued vector $1$-form which adds a spin-$2$, spin-$1$ and a spin-$0$ fields into the reduced theory. The last piece, $\gamma^a$, is a Lorentz-valued vector $0$-form, which adds a new spin-$1$ field into the picture. All these new fields admit the Fourier expansion in Eq.~\eqref{Fourier}, regarding their dependence on the extra dimension. Here, as was pointed out earlier, we will consider only the zero mode of such an expansion.

\subsection{Curvature and torsion}

The $N$-dimensional Lorentz curvature and torsion are given by the Cartan structure equations
\begin{align}
\label{curvadef}
\hat{R}^{AB} &= \mbox{d}\hat{\omega}^{AB}+\hat{\omega}^A_{\ \ C}\wedge\hat{\omega}^{CB}, \\
\label{tordef}
\hat{T}^A &= \hat{\mbox{D}}\hat{e}^A=\mbox{d}\hat{e}^A+\hat{\omega}^A_{\ \ B}\wedge\hat{e}^B,
\end{align}
respectively. Using the definition of curvature in Eq.~\eqref{curvadef}, along with the KK ansatz for the spin connection~\eqref{Nconnection}, we find
\begin{align}\label{R1}
  \hat{R}^{ab}&=R^{ab}+\sqrt{\phi}\alpha^{ab}F-\beta^a\wedge\beta^b\notag\\
  &+\left(\mbox{D}\alpha^{ab}+\frac{1}{2}\alpha^{ab}\mbox{d}\ln\phi-2\beta^{[a}\gamma^{b]}\right)\wedge\hat{e}^N,\\
  \label{R2}
  \hat{R}^{Na}&=-\left(\mbox{D}\beta^a+\sqrt{\phi}\gamma^a F\right)\notag\\
  &+\left(\alpha^a_{\ b}\beta^b-\mbox{D}\gamma^a-\frac{1}{2}\gamma^a\mbox{d}\ln\phi\right)\wedge\hat{e}^N.
\end{align}
where we have used the fact that $\mbox{d}\hat{e}^N=\frac{1}{2}\mbox{d}\ln\phi\wedge\hat{e}^N+\sqrt{\phi}F$. Similarly from the definition of torsion in Eq.~\eqref{tordef}, we find its distinctive parts to be
\begin{align}\label{T1}
  \hat{T}^a &= T^a+\left(\beta^a-\alpha^a_{\ \ b}e^b\right)\wedge\hat{e}^N,\\
  \label{T2}
  \hat{T}^N &= \sqrt{\phi}F-\beta_b\wedge e^b+\left(\frac{1}{2}\mbox{d}\ln\phi+\gamma_be^b\right)\wedge\hat{e}^N.
\end{align}

Considering the Bianchi identities for the KK structure described above, we find relevant information about the new fields. Taking the exterior covariant derivative over the $N$-dimensional curvature and torsion, the Bianchi identities are
\begin{align*}
  \hat{\text{D}}\hat{R}^{AB} &= 0 &\mbox{and}& &\hat{\text{D}}\hat{T}^A &= \hat{R}^A_{\ \ B}\wedge\hat{e}^B.
\end{align*}
A careful decomposition of the first one into its distinctive parts, gives the Bianchi identity for the curvature of the reduced spacetime, together with the second derivative rules
\begin{align*}
  \text{D}R^{ab} &=0, & \text{D}\mbox{D}\alpha^{ab} &=R^a_{\ \ l}\alpha^{lb}+R^b_{\ \ l}\alpha^{al},\\
  \text{D}\text{D}\beta^a &= R^a_{\ \ b}\beta^{b}, & \text{D}\text{D}\gamma^a &= R^a_{\ \ b}\gamma^{b}.
\end{align*}
Considering the second Bianchi identity, gives nothing but its equivalent for the manifold $M_{N-1}$, this is
\begin{equation*}
  \mbox{D}T^a=R^a_{\ \ b}\wedge e^b.
\end{equation*}

This is taken as a proof of the tensorial nature of these new fields under the $4$-dimensional Lorentz transformations, which have been already discussed in Sec.~\ref{KK}.

\section{Five-Dimensional Lovelock--Cartan Reduction\label{5EGB}}

The most general theory requiring the Lagrangian to be (i)~an invariant $N$-form under local Lorentz transformations; (ii)~a local polynomial of the vielbein, the Lorentz connection and their exterior derivatives; (iii)~constructed without the Hodge dual, is the so-called Lovelock--Cartan theory of gravity~\cite{Mardones:1990qc}. This theory is the natural generalization of the Lanczos--Lovelock action, when torsional degrees of freedom are present.

In five dimensions, the action principle of such a theory is given by  
\begin{widetext}
  \begin{equation}\label{action5EGB}
    I=\int\limits_{M_5}\hat{\epsilon}_{ABCDE}\Big(\frac{\alpha_0}{5}\hat{e}^A\wedge\hat{e}^B\wedge\hat{e}^C\wedge
    \hat{e}^D\wedge\hat{e}^E
    +\frac{\alpha_1}{3}\hat{R}^{AB}\wedge\hat{e}^C\wedge\hat{e}^D\wedge\hat{e}^E
    +\alpha_2\hat{R}^{AB}\wedge\hat{R}^{CD}
    \wedge\hat{e}^E\Big),
  \end{equation}
where $\alpha_0$, $\alpha_1$ and $\alpha_2$ are dimensionful coupling constants. Variation with respect to the $5$-dimensional vielbein and spin connection give the equations
  \begin{align}\label{delta e}
    \hat{\epsilon}_{ABCDE}\Big(\alpha_0\hat{e}^A\wedge\hat{e}^B\wedge\hat{e}^C\wedge\hat{e}^D
    +\alpha_1\hat{R}^{AB}\wedge\hat{e}^C\wedge\hat{e}^D
    +\alpha_2\hat{R}^{AB}\wedge\hat{R}^{CD}\Big)&=0,
    \\\label{delta w}
    \hat{\epsilon}_{ABCDE}\Big(\alpha_1\hat{e}^C\wedge\hat{e}^D+
    2\alpha_2\hat{R}^{CD}\Big)\wedge\hat{T}^E&=0,
  \end{align}
respectively.
\end{widetext}

In general, the Lovelock--Cartan theory in five dimensions allows the torsional term~\cite{Mardones:1990qc},
\begin{align}
\label{boundary}
\mathcal{L} \propto \hat{T}_A\wedge \hat{R}^A_{\ B}\wedge\hat{e}^B,
\end{align}
which can be written as a boundary term proportional to the Nieh-Yan invariant, adding no dynamics to the field equations. The action principle~\eqref{action5EGB}, along with the torsional term~\eqref{boundary}, describes the general Lovelock--Cartan theory in five dimensions.

In this work, we will focus on the region in the parameter space where
\begin{equation}\label{delta}
\Delta\equiv\alpha_1^2-4\alpha_0\alpha_2\neq 0.
\end{equation}
This condition place us outside the Chern--Simons point since, otherwise, the field equations~\eqref{delta e} and~\eqref{delta w} remain invariant under $AdS_5$ gauge transformations, while the action principle~\eqref{action5EGB} can be recognized as its Chern--Simons form~\cite{Zanelli:2005sa,Troncoso:1999pk}.
